---
title: "Mini Data-Analysis Deliverable 1"
author: "Kashish Doshi"
date: "October 9th, 2021"
output: github_document
---

# Welcome to your (maybe) first-ever data analysis project!

And hopefully the first of many. Let's get started:

1. Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

```
install.packages("ggbreak")
install.packages("devtools")
devtools::install_github("UBC-MDS/datateachr")



```

2. Load the packages below.

```{r}
suppressMessages(library(datateachr))
suppressMessages(library(tidyverse))
suppressMessages(library(tibble))
suppressMessages(library(viridis))
suppressMessages(library(ggbreak))
```

3. Make a repository in the https://github.com/stat545ubc-2021 Organization. You will be working with this repository for the entire data analysis project. You can either make it public, or make it private and add the TA's and Vincenzo as collaborators. 

4. When you go to submit, submit a URL to your repository to canvas. 

# Instructions

More details regarding the instructions and points allocated to each task can be found below. Briefly,

+ Each milestone is worth 30 points. The number of points allocated to each task will be annotated within each deliverable. Tasks that are more challenging will often be allocated more points.

+ 10 points will be allocated to the reproducibility, cleanliness, and coherence of the analysis. While the three milestones will be submitted as independent deliverables, the analysis itself is a continuum - think of it as 3 chapters to a story. Each chapter, or in this case, portion of your analysis, should be easily followed through by someone unfamiliar with the content. [Here](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/) is a good resource for what constitutes "good code". Learning good coding practices early in your career will save you hassle later on!

# Learning Objectives

By the end of this milestone, you should:

+ Become familiar with your dataset of choosing
+ Think of 4 questions that you would like to answer with your data
+ Generate a reproducible and clear report using R Markdown

# Task 1: Choose your favorite dataset (10 points)

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

+ *apt_buildings*: Acquired courtesy of The City of Toronto’s Open Data Portal. It currently has 3455 rows and 37 columns.

+ *building_permits*: Acquired courtesy of The City of Vancouver’s Open Data Portal. It currently has 20680 rows and 14 columns.

+ *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

+ *flow_sample*: Acquired courtesy of The Government of Canada’s Historical Hydrometric Database. It currently has 218 rows and 7 columns.

+ *parking_meters*: Acquired courtesy of The City of Vancouver’s Open Data Portal. It currently has 10032 rows and 22 columns.

+ *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

+ *vancouver_trees*: Acquired courtesy of The City of Vancouver’s Open Data Portal. It currently has 146611 rows and 20 columns.

**Things to keep in mind**

+ We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

+ Before choosing a dataset, you should always keep in mind **your goal**, or in other ways, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!

1.1 Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

**Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data.

1: _steam_games_  
2: _vancouver_trees_    
3: _parking_meters_    
4: _cancer_sample_ 

1.2 One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*. 

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you, but do you want to use more than one? Would you like to write more comments outside of the code chunk?

__To get the basic attributes of each data set I can use the _str()_ function which gives the number of rows, columns, the names of columns and, the type of variables used.__

### Steam Games

Before working with a data set it is important to understand the types of variables being used here.

```{r}
### Steam Games ###
str(steam_games, 
    stringsAsFactors=FALSE, 
    vec.len = 3, 
    give.length= FALSE, 
    nchar.max=50)

```

There's one interesting column, which is the release date column which is not numeric or date formatted, but rather long strings 

```{r}
steam_games %>%
  arrange(release_date) %>% 
  select(release_date) %>% 
  head(n=8)
```

Interestingly, some of the dates are not even dates but rather *coming soon* or even "¯\\_(ツ)_/¯"

This means date's will have to be edited or even changed before they can be analyzed.This means more pre-processing will have to be done in order for me to analyse the components of this dataset.



### Vancouver Trees

Again, first let's see what kinds of data are present 

```{r}
str(vancouver_trees, 
    stringsAsFactors=FALSE, 
    vec.len = 3, 
    give.length= FALSE, 
    nchar.max=50)
```


The diameter is an interesting column and could be used in multiple plots. I would want to check how many rows are missing observations

```{r}

vancouver_trees %>%
  select(diameter) %>%
  drop_na()%>%
  dim()

```
Fortunately, all the trees have this observation! 

I also wanted to check if the different neighborhoods had similar number of trees sampled in this data set

```{r}

neigh_count = vancouver_trees %>% select(neighbourhood_name) %>%
  table() %>%as_tibble() 
neigh_count_sorted =neigh_count %>% 
  rename(Neighbourhood ="." , Observations = n) %>%
  arrange(desc(Observations))

print(neigh_count_sorted, n=1e2)

```

This table was sorted by the number of observations. Looks like almost all neighborhoods have at least 4000 trees observed. This will make for good analyses. 



### Parking Meters

Again, I will look at the type of data we get from this data set

```{r}
str(parking_meters, 
    stringsAsFactors=FALSE, 
    vec.len = 3, 
    give.length= FALSE, 
    nchar.max=50)
```

The columns have a weird name and I will have to guess what they mean 
I am assuming column 2 which reads __r_mf_9a_6p__ means rate for 1 hour of parking from Monday through to Friday between 9 am and 6pm.

All the rates and hours are tabulated as characters and not numbers. Here again, I will have to extract the numeric value from the columns prior to running any analyses. 
Although this data set has a lot of information, I do not think I want to use it further since I cannot think of a lot of questions I could ask or answer with the given information without heavy cleaning up. 



### Cancer Sample

Let's look at what data is present here.

```{r}
str(cancer_sample, 
    stringsAsFactors=FALSE,
    vec.len = 3, 
    give.length= FALSE, 
    nchar.max=50,
    give.attr=FALSE)

```

Interestingly again the column names are not informative to someone not already familiar with oncology and even then there is no further information given so I would prefer to not use this for my analysis



1.3 Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 2. What lead you to choose these 2? Briefly explain your choices below, and feel free to include any code in your explanation. 

Looking at the outputs above, I narrowed down my choices to _steam_games_ and _vancouver_trees_. 
One of the major reasons is that I can see more potential in these 2 data sets to analyze and be able to ask questions. 
For e.g. __How many games published after 2010 have a overall positive review and how did this change from pre-2010?__

The reason for not selecting the other 2 data sets is that I do not understnad the column names too well and I cannot find more information about them as well as I do not find the data interesting. Especially, in the _cancer_sample_ data-set, the column names do not provide enough information to the average person. Further reasoning for this selection can also be found in the comments provided before in _1.2_.  


1.4 Time for the final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to answer with each dataset. Note them down below, and make your final choice based on what seems more interesting to you!

1. _steam_games_ : Are recent and overall reviews different between games that are free and not free?
2. _vancouver_trees_: Do certain neighborhoods plant more trees of 1 species than other? 
3. _parking_meters_: Are parking meters in certain local areas priced higher on average than parking meters in other areas? 
4. _cancer_sample_ : Are concavity and smoothness positvely co-releated in samples with a "M" diagnosis? 

# Choice of data-set

For my final choice, I find _vancouver_trees_ the most interesting and I feel that it has the most potential to answer questions about people's reviews of recent games, and genres. For further clarification of this choice, please refer to the comments in the previous section. 



# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. Remember, marks will be awarded for completion of the 4 tasks, but 10 points of the whole project are allocated to a reproducible and clean analysis. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset (15 points)

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to answer during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE**.

2.1 Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only answer *4*. Use *dplyr* and *ggplot*.

1. Plot the distribution of a numeric variable.DONE
2. Create a new variable based on other variables in your data (only if it makes sense). DONE
3. Investigate how many missing values there are per variable. Can you find a way to plot this?
4. Explore the relationship between 2 variables in a plot.DONE
5. Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year...DONE. Combined with point 4
6. Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!DONE
7. Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring. Done
8. Use a density plot to explore any of your variables (that are suitable for this type of plot).

2.2 For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

# Task 3: Write your research questions (5 points)

So far, you have chosen a dataset and gotten familiar with it through exploring the data. Now it's time to figure out 4 research questions that you would like to answer with your data! Write the 4 questions and any additional comments at the end of this deliverable. These questions are not necessarily set in stone - TAs will review them and give you feedback; therefore, you may choose to pursue them as they are for the rest of the project, or make modifications!


## Vancouver Trees

First I wanted to look at the distribution of the diameters of all the trees in the data set
```{r}
vancouver_trees %>% 
  ggplot(aes(diameter))+ 
    geom_histogram(bins = 20,na.rm = TRUE, colour = "black")+ 
    xlim(0,60)+
    xlab(label = "Diameter") + 
    ylab(label = "Counts")

```

Here we can see that most of the trees have diameters between 20 and 0. No units were provided so I am _assuming_ they mean _inches_.

It would be interesting to see how the diameter relates to the different species of the trees
First I need to find the average diameter per species. To do this I will create a new coloumn called avg_diameter

```{r}

trees_avg_dia = vancouver_trees %>% 
  group_by(species_name)%>%
  mutate(avg_diameter=mean(diameter, na.rm=TRUE))

trees_avg_dia %>% 
  select(species_name, diameter, avg_diameter)  
```


Then I will plot the species versus their average diameter this could help city planners account for areas for trees in public parks and gardens. In order to fit the plot in the document I will only be looking at top 5 most frequent species. 

```{r}
species_count = as.data.frame(table(trees_avg_dia$species_name)) %>% 
  arrange(desc(Freq))

spec_count = as.tibble(species_count)

species_count_name = spec_count$Var1[1:5]

species_count_name %>% as_tibble() %>% rename(Species=value)

```
From this table we can see the top 5 most frequent species of trees in the data set. 

Next I will filter these species from the data set with the average diameters and also only select 1 row of every species since that is all I need to plot the average diameter. 

```{r}
top_tree_dia = trees_avg_dia %>% 
  select(species_name, avg_diameter) %>% 
  filter(species_name %in% species_count_name)%>% 
  distinct(species_name, .keep_all = TRUE) %>% 
  ggplot(aes(species_name, avg_diameter))+ 
    geom_col(colour="green", fill = "chocolate4")+
    xlab(label = "Species")+ 
    ylab(label = "Average Diameter")

print(top_tree_dia)
```

This plot shows that 4 out of 5 of the most frequently observed tree species have average diameters > 10 inches. 

Another way of showing this is by using box plots
```{r}
vancouver_trees %>%
  filter(species_name %in% species_count_name) %>%
  ggplot(aes(species_name, diameter))+
  geom_boxplot(na.rm=TRUE) + 
  stat_summary(fun = mean,colour= "red", fill="red")+
  scale_y_break(c(70,125), scales="free") + 
  xlab(label = "Species") +
  ylab(label = "Diameter")



```

This plot shows the same information as the last with the means in red. Again 4 out of the 5 most frequently observed tree species have an average diameter > 10 inches. 

I also wanted to check the distribution of tree diameters by neighbouhood. Again to make sure the plot fits the document I have selected the 5 neighbourhoods with the most number of observations. This was defined in the previous section. 

```{r}

vancouver_trees %>%
  group_by(neighbourhood_name) %>%
  filter(neighbourhood_name %in% neigh_count_sorted$Neighbourhood[1:5]) %>% # Used neigh_count_sorted from previous section
  ggplot(aes(neighbourhood_name, diameter, colour=neighbourhood_name))+
    geom_violin(na.rm = TRUE)+
    scale_y_break(c(50,350), scales="free")+ 
    theme(axis.text.x = element_blank(),axis.title.x = element_text(vjust=0.8,hjust = 0.3))+ 
    xlab(label = "Neighbourhood")+
    ylab(label = "Diameter")
  

```


This plot shows that almost all neighborhoods have more trees with diameters <5 inches irrespective of the species of the trees. Dunbar-Southlands seems to have more trees with larger diameters as compared to the other 4. 


### Research Questions

1. Do trees that were planted earlier have larger diameters? Does this change with neighborhoods? 
  In this question I would investigate trees grouped by their species. It would involve comparing the date of trees planted against the diameter and seeing if trees planted earlier than the others have larger diameters. I would also like to see if these species' diameter changes based on the neighborhood
  
2. Were certain species of trees planted more after a certain year? 
  For this analysis I would look at how the overall composition of the different tree species in a year and plot changes over time. This will tell me if a few species start getting planted more or less after a certain year.
  
3. Did the planting of trees in Vancouver have a geographical change over time? 
  Here I would look at the neighborhoods where the trees were planted and check if the trees were planted in all neighborhoods simultaneously or was did only some neighborhoods get trees first and later others were also included. I do understand that the data set does not comprise of each and every tree in Vancouver. But it would be interesting to see if some neighborhoods had "greener thumbs" than others. 
  
4. What is the relationship between the tree diameter and height. Does this change with species? 
  This would involve grouping the tree species and seeing if the diameter and height can be explained via a linear or other models.
  
  
  
  
### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.




























