---
title: "Mini Data Analysis Milestone 2"
output: github_document
---

*To complete this milestone, you can either edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome back to your mini data analysis project!

This time, we will explore more in depth the concept of *tidy data*, and hopefully investigate further into your research questions that you defined in milestone 1.

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

Begin by loading your data and the tidyverse package below: 

```{r, message = FALSE}
suppressMessages(library(datateachr))
suppressMessages(library(tidyverse))
suppressMessages(library(tibble))
suppressMessages(library(viridis))
suppressMessages(library(ggbreak))

```

# Learning Objectives

By the end of this milestone, you should:

+ Become familiar with manipulating and summarizing your data in tibbles using `dplyr` and `tidyr`, with a research question in mind.
+ Understand what *tidy* data is, and how to create it. In milestone 3, we will explore when this might be useful.
+ Generate a reproducible and clear report using R Markdown.
+ Gain a greater understanding of how to use R to answer research questions about your data.

**Things to keep in mind**

+ Remember to document your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. Create your analysis as if someone else will be reading it! **There will be 2.5 points reserved for reproducibility, readability, and repo organization.**

+ Before working on each task, you should always keep in mind the specific **research question** that you're trying to answer.

# Task 1: Process and summarize your data (15 points)

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (2.5 points) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->
1.  Do trees that were planted earlier have larger diameters?

2.  Were certain species of trees planted more after a certain year? 

3.  Did the planting of trees in Vancouver have a geographical change over time? 

4.  What is the relationship between the tree diameter and height? 
<!----------------------------------------------------------------------------->

### 1.2 (10 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question. 

Ensure that the output of each operation is printed! 

**Summarizing:**

1. Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2. Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3. Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Based on two categorical variables, calculate two summary statistics of your choosing.

**Graphing:**

5. Create a graph out of summarized variables that has at least two geom layers.
6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read. 
7. Make a graph where it makes sense to customize the alpha transparency.
8. Create 3 histograms out of summarized variables, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->
### Research Question 1: Do trees that were planted earlier have larger diameters?

To evaluate this question, I can start by extracting the years from the _date_planted_ column and categorizing the data by treating teh years as factors. Next, I calculate the mean diameters for each year.  
```{r}
trees_time_planted = vancouver_trees %>% 
  na.omit(date_planted) %>%
  mutate(year_planted=format(date_planted, format="%Y")) %>% 
  mutate(year_planted = factor(as.integer(year_planted))) %>% 
  arrange(year_planted) %>%
  group_by(year_planted) %>% 
  mutate(mean_dia_yr = mean(diameter))

print(trees_time_planted)
#Summarizing 3. answered
```


Having all the data ready to plot. I can use a a point plot in ggplot to display the diameters per year and overlay the mean diameters as a line to check for a relationship.
```{r}
plot_TimeVsDia =trees_time_planted %>% 
  ggplot()+
  geom_point(aes(year_planted,diameter), alpha = 0.5, colour = "cyan4") + 
  geom_line(aes(year_planted, mean_dia_yr, group = 1)) +
  geom_point(aes(year_planted, mean_dia_yr), colour = "red")+
  ylim(... = c(0, 50))+
  theme(axis.text.x = element_text(angle = 90)) + 
  xlab(label = "Year Planted") +
  ylab(label = "Diameter")+
  labs(title = "Relationship between date planted and diameter")

print(plot_TimeVsDia)
#Graphing 5 and 7 answered
```

Here, the black line is the mean diameter of trees over the years and the red points are the values of the mean diameter for each year.
Note: I removed 11 observations contianing Outliers to improve the resolution of the plot.

From the plot, there seems to be a small decrease in mean diameter as the year of planting increases. This seems obvious as trees that were planted later, would be younger and thus would be smaller. However, looking at the overall diameters, the range of diameters seen is pretty much identical and so the decrease in mean diameters does not seem meaningful. 


### Research Question 2: Were certain species of trees planted more after a certain year?

To look at this question first I need to modify the data table. First again, I will extract the years from the _date_planted_ column. Next, I will count the number of unique species and add them to a separate column called _n_spec_. Finally, I would also like to see the increase/decrease of species in the current year as compared to the previous year. To this I will create a new column called _n_spec_prev_year_ which will contain values of _n_spec_ from the previous year. Finally to calculate the change in number of species I will subtract _n_spec_prev_year_ from _nspec_ for each year.  

```{r}

species_num = vancouver_trees %>% 
  select(species_name, date_planted) %>% 
  na.omit(date_planted) %>%
  mutate(year_planted=as.integer(format(date_planted, format="%Y"))) %>% 
  group_by(year_planted) %>% 
  arrange(species_name, year_planted) %>%
  mutate(n_spec= length(unique(species_name))) %>% 
  arrange(year_planted) %>% 
  distinct(n_spec, .keep_all = TRUE) %>% 
  ungroup(year_planted) %>%
  mutate(n_spec_prev_year = append(0, n_spec[-length(n_spec)]), diff_spec = n_spec-n_spec_prev_year)

print(species_num)
#Summarizing 2 answered
```



To graph the date I will create a line plot which will show the number of species planted each year. This will be overlaid by a column plot that will show the increase/decrease of species over each year. 

```{r}

plot_SpecVsTime = species_num %>% 
  group_by(year_planted)%>% 
  ggplot() + geom_col(aes(year_planted, diff_spec), alpha = 0.5, fill="cyan4")+ 
  geom_line(aes(year_planted, n_spec))+
  scale_y_continuous(name = "Number of Species")+
  xlab(label = "Year Planted")+
  labs(title = "Number of species planted over time")

print(plot_SpecVsTime)
#Graphing 5 answered 

```
Interestingly, from this plot we can see that in 1990 there were about 7 new species planted and 2003 was the year when most species were planted in Vancouver. 

This data shows that there was a change in the species of trees planted from 1990 to 2019.  



### Research Question 3: Did the planting of trees in Vancouver have a geographical change over time?

To look at the geographical change I will be looking at the neighborhoods where the trees planted were recorded. First, I will extract the years of the _date_planted_. Next I will count the number of trees planted per year in each neighborhood and put them in a new column called _n_planted_. I also looked at the average number of trees planted per year and put this data in the column _avg_tree_pl_year_.
```{r}

trees_neigh = vancouver_trees %>% 
  na.omit(date_planted) %>%
  mutate(year_planted=as.integer(format(date_planted, format="%Y"))) %>%
  group_by(neighbourhood_name, year_planted) %>%
  mutate(n_planted = sum(complete.cases(year_planted))) %>%
  ungroup(neighbourhood_name) %>%
  mutate(avg_tree_pl_year = mean(n_planted)) %>% group_by(neighbourhood_name)

print(trees_neigh)
# Summarizing 2 asnwered
```


In order to make informative yet not messy graphs I decided to look at the 5 neighborhoods with the most number of trees planted. To look at this, I extract the relevant columns from the previous data. Next, I count the total number of trees recorded per neighbourhood by look at the number of observations in _year_planted_ column per neighborhood. Next the data is sorted by this count to obtain the neighbourhoods with the most number of trees. I selected the first 5. 
```{r}
most_pop_neigh = trees_neigh %>%
  arrange(neighbourhood_name, year_planted,n_planted) %>% 
  select(neighbourhood_name, year_planted, n_planted) %>% 
  unique() %>% ungroup(year_planted) %>% 
  mutate(total_planted = sum(n_planted)) %>% 
  distinct(total_planted,.keep_all = TRUE) %>% 
  arrange(desc(total_planted)) %>% select(neighbourhood_name, total_planted)

head(most_pop_neigh, n=5)
```

Finally to plot the data, I first filter the  data to only keep the neighborhoods mentioned above. I created a line plot to show the number of trees planted per year and grouped them by neighborhood. I also added a second line plot with the average number of trees planted per year to see if there was 1 neighbourhood that would be responsible to drive the average up. 
```{r}
plot_neigh_time = trees_neigh %>% 
  filter(neighbourhood_name %in% most_pop_neigh$neighbourhood_name[1:5]) %>% 
  group_by(neighbourhood_name) %>% 
  ggplot(aes(colour = neighbourhood_name))+
  geom_line(aes(year_planted, n_planted, group = neighbourhood_name), alpha = 0.9) + 
  geom_point(aes(year_planted, n_planted), alpha = 0.3) + 
  geom_line(aes(year_planted, avg_tree_pl_year), colour = "black") + 
  xlab(label = "Year Planted") + 
  ylab(label = "Number of Trees Planted")+
  labs(title = "Trees planted by year in each neighbourhood", colour= "Neighbourhood Name")

print(plot_neigh_time)
#Graphing 5 and 7 answered
```
The black line in the plot above shows the average number of trees planted per year. Although the data looks messy, the information from this graph is very simple. It seems like all neighbourhoods planted an similar number of trees over the years and no neighbourhood particularly stands out. However some information obtained from this graph shows that Kensington-Cedar Cottage planted its highest number of trees in 2001 and this was the highest in my data set too!. Similarly Renfrew-Collingwood reached its planting peak in 2007.  

Overall it does not seem like planting of trees in Vancouver had a geographical change.



### Research Question 4: What is the relationship between the tree diameter and height?

Too answer this question I decided only to look at teh genus of trees since I feel that on a species level, trees of the same genus may have similar diameters and heights. I first calculated the mean diameter and height for each genus. To look at more statistics I also calculated the minimum, maximum and range of diameters and heights for each genus and put them into relevant columns.  

```{r}
trees_dia_vs_height = vancouver_trees %>% 
  select(genus_name, diameter, height_range_id) %>% 
  group_by(genus_name) %>% 
  mutate(dia_mean = mean(diameter, na.rm=TRUE),
         dia_max = max(diameter),
         dia_min = min(diameter),
         dia_range = (dia_max - dia_min),
         height_mean = mean(height_range_id, na.rm=TRUE),
         height_max = max(height_range_id),
         height_min = min(height_range_id),
         height_range = height_max - height_min)

print(trees_dia_vs_height)
#Summarizing 1 answered

```


In order to prevent messy and complicated plots I decided to only look at the 8 genus with the most number of trees plotted. To obtain these genus, I counted the number of trees in each genus and sorted this data by the number of observations. 
```{r}
pop_genus = table(trees_dia_vs_height$genus_name) %>% 
  as_tibble(.name_repair = "minimal") %>% 
  rename(Genus_name = "") %>% 
  arrange(desc(n))

head(pop_genus, n=8)
```

To plot the data, I filtered the original dataset to only contain rows with the genus mentioned above. Next I used a dot plot to plot average diameters against heights of the trees. I also highlighted the minimum and maximum heights in each group. Finally I overlaid a line plot showing the mean heights which would show the relationship between the height and diameter. 

```{r}
plot_DiaVsHeight = trees_dia_vs_height %>% 
  filter(genus_name %in% pop_genus$Genus_name[1:8]) %>% 
  arrange(dia_mean) %>% 
  ggplot(aes(dia_mean, height_range_id))+
  geom_point(aes(colour = genus_name), alpha= 0.3) + 
  geom_line(aes(dia_mean, height_mean)) + 
  geom_point(aes(dia_mean, height_max), colour = "red", pch=21) + 
  geom_point(aes(dia_mean, height_min), colour = "black", pch=21) + 
  xlab(label = "Average Diameter") + 
  ylab(label = "Height Range")+
  labs(title = "Relationship between tree diameter and height", colour = "Genus Name")

print(plot_DiaVsHeight)
#Graphing 5 and 7 answered
```
In the plot above the minimum and maximum observations for each genus are shown in black and red bordered dots respectively. The black line shows the mean diameters according to the mean heights of the genus. 

From this data it does not seem like there is a relationship between height and diameter since trees of the genus _Prunus_ have the second highest mean diameters but do not have the second highest mean heights. Perhaps including all genus recorded would lead to a more evident relationship.  



<!----------------------------------------------------------------------------->

### 1.3 (2.5 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->
1.  Do trees that were planted earlier have larger diameters?
Although the data showed that trees planted earlier did have larger diameters, the difference was quite small to be conclusive. This can be seen in the large range of diameters seen every year which brought mean diameters for each year close to each other. Perhaps the certain genus or species could show a more conclusive relationship. A better question to ask would be:

__"Does the year of planting have an effect on the diameter on a species-specific basis?"__

2.  Were certain species of trees planted more after a certain year? 
According to the graphs it seems like there was a change in the number of species planted every year and therefore requires more analysis to lead to a more conclusive answer. The analysis performed only showed that there was a difference in the number of species planted over the years but not the specific species. Here I feel that I am closer to the answer but not completely there yet. A more refined question in this case would be:

__"When did certain species of trees start getting planted and were these stopped being planted after a certain year?"__

3.  Did the planting of trees in Vancouver have a geographical change
    over time? 
After looking at the data generated it seems that there isn't a difference in the number of trees planted per year in the neighborhoods of Vancouver and that each neighborhood planted similar number of trees overall. However, my analysis only looked at 5 neighbourhoods and there could be a difference if I included all the neighbourhoods. Perhaps there is a difference in which species were planted in each neighborhood over the years. A more specific question to ask would be: 

__"Did some neighborhoods start planting trees of the genus _Acer_ more after 1989?"__

4.  What is the relationship between the tree diameter and height? 
From the data it seems that there is no meaningful relationship between height and diameter of the trees. However, this data was limited to the 5 most planted genus and there may be a relationship if all the genus were taken into account. 

<!----------------------------------------------------------------------------->

# Task 2: Tidy your data (12.5 points)

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

- Each row is an **observation**
- Each column is a **variable**
- Each cell is a **value**

*Tidy'ing* data is sometimes necessary because it can simplify computation. Other times it can be nice to organize data so that it can be easier to understand when read manually. 

### 2.1 (2.5 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have >8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->
```{r}
vancouver_trees %>% select(std_street, genus_name, species_name, height_range_id, on_street,date_planted, longitude, latitude)

```

Based on the table above, I think _genus_name_, _species_name_, __on_street_block__, _height_range_id_, _longitude_ and, _latitude_ are all examples of tidy data since each column is 1 variable and each cell is 1 value of that variable. But there are also instances of untidy data, for example:

_date_planted_ : This column contains data in the date format but is not tidy for analysis. I had to extract the year from the column every time I wanted to run any time based analysis in my work and it should have been divided into day, month and year.

_std_street_ : Here again it would be tidy'er to have separated the street name and the street type to make for better analysis.

<!----------------------------------------------------------------------------->

### 2.2 (5 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->

Starting with the data seen above we can look at the data "as is":
```{r}
untidy_data = vancouver_trees %>% 
  select(std_street, genus_name, species_name, height_range_id, on_street,date_planted, longitude, latitude)

print(untidy_data)
```


Next I will separate the data_planted into columns "year", "month" and, "day". Separating this column will help to run analyses by year or even round the years into decades and run the analyses by decade which was not possible earlier.  
I will also separate std_street into "Street Direction", "Street Name" and, "Street Type". This again will help to run analyses answering questions such as "Do trees planted on streets facing East reach higher diameters?".
```{r}
tidy_data = untidy_data %>% 
  separate(date_planted, into = c("year", "month", "day"), sep = "-") %>%
  separate(std_street, into = c("Street Direction","Street Name", "Street Type"), sep = " ", fill = "left")

print(tidy_data)
```


Finally, I will untidy the data back to what it was
```{r}
untidied_data = tidy_data %>%
  unite(col="year_planted", c(year, month, day), sep="-", na.rm=FALSE) %>%
  unite(col="std_street", c(`Street Direction`,`Street Name`, `Street Type`), sep = " ", na.rm = TRUE)

print(untidied_data)

```

<!----------------------------------------------------------------------------->

### 2.3 (5 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in milestone 3, and explain your decision. 

Try to choose a version of your data that you think will be appropriate to answer these 2 questions in milestone 3. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelvant columns, etc.). 

<!--------------------------- Start your work below --------------------------->
After thinking about the data more, I have decided to continue working on research questions 1 and 2:

1.  Do trees that were planted earlier have larger diameters?
I want to answer this question because I feel that grouping the data according to genus or species has potential to show a relationship between year planted and diameters.I am interested in this question and I would like to pursue this to the end. There is also a possibility that I could find a relationship such that after a certain year there would no affect on the diameter notwithstanding the age of the tree. 

2.  Were certain species of trees planted more after a certain year?
I feel that my analysis here was almost complete and only requires refining as well a bit of modifying. This question still remains unanswered and I am interested in the results which could show trends in tree planting of the residents of Vancouver. 

For the next set of analyses I will tidy _vancouver_trees_ and clean it up so that it is more usable downstream

```{r}
tidy_trees = vancouver_trees %>% 
  filter(!is.na(date_planted)) %>%
  separate(date_planted, into = c("year","month", "day"), sep = "-") %>%
  mutate(year = as.integer(year), month = as.integer(month), day = as.integer(day))%>%
  rename(genus = genus_name, species = species_name, neighbourhood = neighbourhood_name, height_range = height_range_id) %>%
  select(genus, species, neighbourhood, diameter, height_range, year)

print(tidy_trees)
```

<!----------------------------------------------------------------------------->

*When you are done, knit an `md` file. This is what we will mark! Make sure to open it and check that everything has knitted correctly before submitting your tagged release.*

### Attribution

Thanks to Victor Yuan for mostly putting this together. 