---
title: "Mini Data Analysis 1"
author: "Kashish Doshi"
date: "05/10/2021"
output: html_document
---
# Milestone 1 

The goal of this project is to explore and analyze a data set from the datateachr package.

### Installing and Loading the necessary packages
```{r setup}
#install.packages("devtools")
#library(devtools)
#devtools::install_github("UBC-MDS/datateachr")
library(datateachr)
library(tidyverse)
```
### 1.1 Selecting datasets

4 data sets from the datateachr package were selected

#### 1. Steam Games

First let's look at the dimensions of the data set
```{r}
dim(steam_games)
```
Shows that __steam_games__ has 40,833 rows and 21 columns

Next, let's look at the all the types of columns present 
```{r}
str(steam_games, stringsAsFactors=FALSE, vec.len = 3, give.length= FALSE, nchar.max=20)
```

Interestingly, The release dates for the games are not numeric. I would like to see if I can still arrange the data set according to the release dates
```{r}
steam_games %>% arrange(release_date) %>% head(n=5)
```
Seems like some of the dates are not even dates. If this data set is used ahead, these values may need to be edited. 


#### 2 Vancouver Trees
Again let's look at the dimensions of the data set
```{r}
dim(vancouver_trees)
```
Then, extract the different column names and types
```{r}
str(vancouver_trees, stringsAsFactors=FALSE, vec.len = 3, give.length= FALSE, nchar.max=20)

```
The diameter seems like an important variable for plotting and analysis so let's see how many trees do not have a diameter recorded 
```{r}
drop_na(vancouver_trees, diameter) %>% dim()

```
Fortunately, all trees have a diameter recorded!

Maybe the Latitude and Longitude of the data set could be defined by a certain range to see if all the tree samples are neighboring each other or not.
```{r}
lat = vancouver_trees %>% select(latitude) %>% range(na.rm= TRUE) %>% as.character()
cat("Latitudinal Range ", lat[1], " ", lat[2])
long = vancouver_trees %>% select(longitude) %>% range(na.rm=TRUE)
cat("Longitudinal Range ", long[1], " ", long[2])
```
Seems like all the trees in this data set are neighbors! 

#### 3  Parking Meters

Let's look at the dimensions of the data set
```{r}
dim(parking_meters)

```
There are 10032 rows and 22 columns

Let's look at the type of data in this data set
```{r}
str(parking_meters, stringsAsFactors=FALSE, vec.len = 3, give.length= FALSE, nchar.max=20)

```
The columns have a weird name and I will have to guess what they mean 
I am assuming column 2 which reads __r_mf_9a_6p__ means rate for 1 hour of parking from Monday through to Friday between 9 am and 6pm




#### 4 Cancer Sample

```{r}
cancer_sample


```
## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
